\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-04 paper/main.tex
refactoring intro

A core requirement of modern scientific research is replicability.
For computational research, replicability means that code should
produce the same results, even when run on different systems.  The
standard solution to improving replicability entails packaging a
project's dependencies along with its primary code base.  Existing
solutions vary in how deeply these dependencies are specified, ranging
from virtual environments (which specify all Python package versions),
to containers (which also specify the operating system), to
virtual machines (which also specify hardware layers of the system).
Each of these existing solutions requires installing or setting up a
system for running the desired code that must be packaged alongside
the primary code base.  Here we propose an even lighter-weight
solution than virtual environments: the \texttt{davos} library.  When
used in combination with a notebook-based Python project, \texttt{davos}
library provides a mechanism for specifying (and automatically
installing) the correct package versions of the project's.  This
enables researchers to share a complete reproducible environment using
a single iPython notebook file.

@@

\begin{table}[H]
\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Notes} \\
\hline
C1 & Current code version &  v0.1.1 \\
\hline
C2 & Permanent link to code/repository used for this code version & \url{https://github.com/ContextLab/davos/tree/v0.1.1} \\
\hline
C3 & Code Ocean compute capsule & \\
\hline
C4 & Legal Code License & MIT \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used & Python, JavaScript, PyPI/pip, IPython, Jupyter, Ipykernel, PyZMQ. Additional tools used for tests: pytest, Selenium, Requests, mypy, GitHub Actions \\
\hline
C7 & Compilation requirements, operating environments \& dependencies & Dependencies:~Python$>$=3.6, packaging, setuptools.~Supported OSes: MacOS, Linux, Unix-like.~Supported IPython environments: Jupyter notebooks, JupyterLab, Google Colaboratory, Binder, IDE-based notebook editors. \\
\hline
C8 & Link to developer documentation/manual & \url{https://github.com/ContextLab/davos\#readme} \\
\hline
C9 & Support email for questions & contextualdynamics@gmail.com \\
\hline
\end{tabular}
\caption{Code metadata}
\label{}
\end{table}

@@

Code sharing is a core component of the open science movement
that has inspired a full ecosystem of tools and packages.  However,
sharing code, in and of itself, does not guarantee that others will be
able to \textit{reproduce} the desired results.  For eample, research
code often requires installing other software packages that extends
the language's basic functionality.  Within the Python community,
external packages that are published in the most popular 
repositories~\citep{PIP, CONDA} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despire that it is \textit{possible} to manually install the intended version numbers of
every dependency of a Python script or package, doing so may cause
conflicts within the user's computing environment that interferes with
the functionality of  \textit{other} code.

@@

At another extreme, virtual machines~citep{VM} provide a complete hardware-level
simulation of the desired system.  Virtual machines are typically
fully isolated from the user's system such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.  Containers~\citep[e.g.,][]{Merk14,
  SINGULARITY} provide a similar ``isolated'' experience.  Although
containerized environments do not specify hardware-level operations,
they are typically packaged with a complete operating system, in
addition to a complete copy of Python and any relevant dependencies.
Virtual environments~\citep[e.g.,][]{vanRoEtal14} also provide a
computing environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproducee a complete copy of the original operating system.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware-level, operating system
level, and Python environment, respectively) that the relevant code
will run similarly for different users.  However, each of these
systems also requires installing independent tools that
can be resource intensive or burdensome to install or configure.

@@

% STOPPED HERE...

Sharing a project's code base is most valuable when that code may be
run on

Modern scientific research frequently entails writing software code
for a wide variety of purposes throughout the scientific process.
Researchers across disciplines may design and implement complex
experiments; collect, store, and analyze large datasets; create
visualizations for presentations and publications; and share their
findings and techniques with peers, students, and the broader public
through tutorials, demos, workshops, and classes.  However, one
fundamental requirement of virtually any form of research-related code
is that its behavior and outputs remain consistent and predictable, no
matter when, where, or by whom it is run.  This stability can be
crucial, for example, to ensuring that data are collected under the
same conditions (e.g., across recordings, subjects, or physical
locations)
%^ or locations, e.g., in different test rooms, on different machines, at different sites, etc.
over multiple months or years, and that they can be accessed,
processed, and analyzed consistently by a research team that may be
spread across multiple institutions or countries.  Additionally,
modern open science practices encourage publicly sharing research code
and data so that others may explore, reproduce, learn from, and build
upon existing work.  Much of the benefit afforded by freely available
research code depends on users' ability to execute it and achieve the
same result as its original author.

Python \cite{vanR95} has become one of the most widely used and
fastest-growing scientific programming languages, in part by combining
an accessible, high-level syntax with a rich ecosystem of powerful
third-party tools that facilitate rapid development and collaboration
\cite{MullEtal15}.  The Python ecosystem offers an extensive data
science toolkit with platforms for interactive programming (e.g.,
Project Jupyter \cite{KluyEtal16}, Google Colaboratory),
community-maintained libraries for data manipulation (e.g., NumPy
\cite{HarrEtal20}, SciPy \cite{VirtEtal20}, Pandas \cite{McKi10}) and
visualization (e.g., Matplotlib \cite{Hunt07}, seaborn \cite{Wask21}),
frameworks for training complex machine learning models (e.g.,
scikit-learn \cite{PedrEtal11}, TensorFlow \cite{AbadEtal15}, Hugging
Face \cite{WolfEtal20}), and myriad other resources.  However, this
heavy emphasis on third-party libraries also presents a challenge to
writing and sharing stable, reproducible scientific Python code:
different versions of the same library may behave differently, adopt
changes in syntax, expose different functions and interfaces, add or
drop support for specific hardware or software, write (or expect to
read) files in different formats, fix (or introduce) bugs, and so on.
While these issues exist to some extent in any software language or
ecosystem, they have a particular impact on the Python community due
to its unusually rapid growth relative to other languages.  Ensuring
Python code behaves consistently over time and across users therefore
typically requires ensuring it is always run with the same specific
set of versions for each third-party package used.
%Ensuring Python code behaves consistently over time and across users therefore typically requires guaranteeing that the same specific version of each required third-party package is used, every time it is run.

One common approach to solving this problem is to create containerized
or virtualized Python environments (e.g., using Docker \cite{Merk14},
Singularity \cite{KurtEtal17}, or conda \cite{Anac12}) tailored to
individual applications.  A researcher may build such an environment
around a particular experiment or analysis pipeline, and exclusively
run their code from inside it, ``entering" and ``exiting" the
environment before and after each time they do so.  They can also
distribute their custom environment alongside their code as a
configuration file that explicitly lists required package versions,
enabling others to build identical copies for themselves.  This allows
research teams to deploy experiments on multiple machines for more
efficient data collection, collaborate on analyses without introducing
conflicts or inconsistencies, and publicly share their study designs
and results for others to reproduce, replicate, or adapt to study new
questions in the future.

While often effective, this approach bears two notable drawbacks.
First, it can add substantially to the technical knowledge, computing
resources, and initial setup needed to run or share the actual code of
interest.  For example, sharing code for an analysis or tutorial that
relies on a particular Docker image to run properly would of course
necessitate writing and distributing extra configuration files and
setup instructions.  But far more burdensomely, it also requires that
anyone who may want to run the code (in addition to the author seeking
to share it!) first be able to install and navigate additional
software that is likely far more complex and resource-intensive than
the actual analysis or tutorial code it facilitates running.  This can
introduce a need for both a degree of computer literacy and
computational resources that may not be universally accessible,
particularly to students or other early-career scientists hoping to
learn from publicly available tutorials.  These added prerequisites
clash with the simplicity and accessibility that have contributed to
Python's popularity, and can create significant barriers to both
contributing to and taking advantage of open science resources.

Second, while many existing tools allow users to initially populate a
Python environment with a fixed set of packages and package versions
(e.g., from a \texttt{requirements.txt}, \texttt{pyproject.toml},
\texttt{environment.yml}, \texttt{Pipfile}, \texttt{Dockerfile RUN}
instruction, etc.), few, if any, ensure that these specified
requirements \textit{remain} satisfied after they are first installed.
The ability to modify an environment after its creation is useful in
many cases (e.g., to install additional software, when
needed). However, this also makes it easy to inadvertently alter
existing packages, potentially leading to subtle issues with code that
relies on them.  For instance, suppose a researcher has implemented a
series of analyses using version 1.0 of ``Package \textit{X}," and
decides to perform an additional analysis that requires installing
``Package \textit{Y}."  If Package \textit{Y} depends on version 0.9
of Package \textit{X}, then Package \textit{X} will be downgraded to
accommodate this new requirement, potentially altering or breaking
previous analyses when they are rerun later, either by the researcher
or someone with whom they've shared their code.  Further, if some
analyses require Package \textit{Y} while others rely on features of
Package \textit{X} not implemented until version 1.0, it's unclear
which version the researcher should install in their environment.

%The \texttt{davos} package provides a novel, Python-native framework for creating, sharing, and running reproducible workflows that was designed to address each of these issues.
The \texttt{davos} package provides a novel, Python-native framework
for creating reproducible workflows that was designed to address each
of these issues. \texttt{davos} allows users to specify dependencies
directly within the code that...

\comment {
POINTS TO MAKE/THINGS TO SAY HERE TO WRAP UP (or maybe save some for impact section?):
- davos simplifies the process of writing, sharing, and running reproducible code (eliminates need for additional complex env management tools, syntax is simple, packages are just installed at runtime without additional setup beforehand, etc.)...
- ...it also provides stability/additional protections against accidental changes in ways other tools don't (i.e., addresses issue 2 above) by verifying (and installing if necessary) requirements each time code is run
- ... while simultaneously enabling new package management patterns (using multiple versions versions of package in same notebook, etc.)
- it's lightweight and intuitive (unlike big/complex env management tools like docker, etc.)
- "davos was specifically designed for use with Jupyter notebook environments, and can be used either on its own or in tandem with another environment management tool"
- "\texttt{davos} was designed specifically for use in Jupyter notebook (formerly called IPython notebook \cite{PereGran07}) environments, and supports Jupyter notebooks \cite{KluyEtal16}, JupyterLab \cite{GranGrou16}, Google Colaboratory, Binder \cite{RagaWill18}, and IDE-based notebook editors."

- maybe don't mention smuggle statement \& onion comment explicitly in this paragraph since they haven't been defined/explained yet, or if we do, then add some sort of "see \_\_\_ below" with \\ref to relevant subsection
}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{snippets/snippet1.pdf}
\caption{\small \textbf{Figure 1}}
\label{fig:snippet1}
\end{figure}

>>>>

A core requirement of modern scientific research is replicability.
For computational research, replicability means that code should
produce the same results, even when run on different systems.  The
standard solution to improving replicability entails packaging a
project's dependencies along with its primary code base.  Existing
solutions vary in how deeply these dependencies are specified, ranging
from virtual environments (which specify all Python package versions),
to containers (which also specify the operating system), to
virtual machines (which also specify hardware layers of the system).
Each of these existing solutions requires installing or setting up a
system for running the desired code that must be packaged alongside
the primary code base.  Here we propose an even lighter-weight
solution than virtual environments: the \texttt{davos} library.  When
used in combination with a notebook-based Python project, \texttt{davos}
library provides a mechanism for specifying (and automatically
installing) the correct package versions of the project's.  This
enables researchers to share a complete reproducible environment using
a single Jupyter notebook file.

@@

\begin{table}[H]
\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata value} \\
\hline
C1 & Current code version &  v0.1.1 \\
\hline
C2 & Permanent link to code/repository used for this code version & \url{https://github.com/ContextLab/davos/tree/v0.1.1} \\
\hline
C3 & Code Ocean compute capsule & \\
\hline
C4 & Legal Code License & MIT \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used & Python, JavaScript, PyPI/pip, IPython, Jupyter, Ipykernel, PyZMQ. Additional tools used for tests: pytest, Selenium, Requests, mypy, GitHub Actions \\
\hline
C7 & Compilation requirements, operating environments \& dependencies & Dependencies:~Python$>$=3.6, packaging, setuptools.~Supported OSes: MacOS, Linux, Unix-like.~Supported IPython environments: Jupyter notebooks, JupyterLab, Google Colaboratory, Binder, IDE-based notebook editors. \\
\hline
C8 & Link to developer documentation/manual & \url{https://github.com/ContextLab/davos\#readme} \\
\hline
C9 & Support email for questions & contextualdynamics@gmail.com \\
\hline
\end{tabular}
\caption{Code metadata}
\label{}
\end{table}

@@

Code sharing is a core component of the open science movement
that has inspired a full ecosystem of tools and packages.  However,
sharing code, in and of itself, does not guarantee that others will be
able to \textit{reproduce} the desired results.  For eample, research
code often requires installing other software packages that extends
the language's basic functionality.  Within the Python community~\citep{vanR95},
external packages that are published in the most popular 
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despire that it is \textit{possible} to manually install the intended version numbers of
every dependency of a Python script or package, doing so may cause
conflicts within the user's computing environment that interferes with
the functionality of  \textit{other} code.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python ecosystem.}}
\label{fig:code-sharing}
\end{figure}

@@

At another extreme, virtual machines~\citep{Gold74, AltiEtal05, vmware} provide a complete hardware-level
simulation of the desired system.  Virtual machines are typically
fully isolated from the user's system such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.  Containers~\citep[e.g.,][]{Merk14,
  KurtEtal17} provide a similar ``isolated'' experience.  Although
containerized environments do not specify hardware-level operations,
they are typically packaged with a complete operating system, in
addition to a complete copy of Python and any relevant dependencies.
Virtual environments~\citep[e.g.,][]{vanREtal14} also provide a
computing environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproducee a complete copy of the original operating system.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware-level, operating system
level, and Python environment, respectively) that the relevant code
will run similarly for different users.  However, each of these
systems also requires installing independent tools that
can be resource intensive or burdensome to install or configure.

@@

% % STOPPED HERE...

% Sharing a project's code base is most valuable when that code may be
% run on

% Modern scientific research frequently entails writing software code
% for a wide variety of purposes throughout the scientific process.
% Researchers across disciplines may design and implement complex
% experiments; collect, store, and analyze large datasets; create
% visualizations for presentations and publications; and share their
% findings and techniques with peers, students, and the broader public
% through tutorials, demos, workshops, and classes.  However, one
% fundamental requirement of virtually any form of research-related code
% is that its behavior and outputs remain consistent and predictable, no
% matter when, where, or by whom it is run.  This stability can be
% crucial, for example, to ensuring that data are collected under the
% same conditions (e.g., across recordings, subjects, or physical
% locations)
% %^ or locations, e.g., in different test rooms, on different machines, at different sites, etc.
% over multiple months or years, and that they can be accessed,
% processed, and analyzed consistently by a research team that may be
% spread across multiple institutions or countries.  Additionally,
% modern open science practices encourage publicly sharing research code
% and data so that others may explore, reproduce, learn from, and build
% upon existing work.  Much of the benefit afforded by freely available
% research code depends on users' ability to execute it and achieve the
% same result as its original author.

% Python \cite{vanR95} has become one of the most widely used and
% fastest-growing scientific programming languages, in part by combining
% an accessible, high-level syntax with a rich ecosystem of powerful
% third-party tools that facilitate rapid development and collaboration
% \cite{MullEtal15}.  The Python ecosystem offers an extensive data
% science toolkit with platforms for interactive programming (e.g.,
% Project Jupyter \cite{KluyEtal16}, Google Colaboratory),
% community-maintained libraries for data manipulation (e.g., NumPy
% \cite{HarrEtal20}, SciPy \cite{VirtEtal20}, Pandas \cite{McKi10}) and
% visualization (e.g., Matplotlib \cite{Hunt07}, seaborn \cite{Wask21}),
% frameworks for training complex machine learning models (e.g.,
% scikit-learn \cite{PedrEtal11}, TensorFlow \cite{AbadEtal15}, Hugging
% Face~\cite{WolfEtal20}), and myriad other resources.  However, this
% heavy emphasis on third-party libraries also presents a challenge to
% writing and sharing stable, reproducible scientific Python code:
% different versions of the same library may behave differently, adopt
% changes in syntax, expose different functions and interfaces, add or
% drop support for specific hardware or software, write (or expect to
% read) files in different formats, fix (or introduce) bugs, and so on.
% While these issues exist to some extent in any software language or
% ecosystem, they have a particular impact on the Python community due
% to its unusually rapid growth relative to other languages.  Ensuring
% Python code behaves consistently over time and across users therefore
% typically requires ensuring it is always run with the same specific
% set of versions for each third-party package used.
% %Ensuring Python code behaves consistently over time and across users therefore typically requires guaranteeing that the same specific version of each required third-party package is used, every time it is run.

% One common approach to solving this problem is to create containerized
% or virtualized Python environments (e.g., using Docker \cite{Merk14},
% Singularity \cite{KurtEtal17}, or conda \cite{Anac12}) tailored to
% individual applications.  A researcher may build such an environment
% around a particular experiment or analysis pipeline, and exclusively
% run their code from inside it, ``entering" and ``exiting" the
% environment before and after each time they do so.  They can also
% distribute their custom environment alongside their code as a
% configuration file that explicitly lists required package versions,
% enabling others to build identical copies for themselves.  This allows
% research teams to deploy experiments on multiple machines for more
% efficient data collection, collaborate on analyses without introducing
% conflicts or inconsistencies, and publicly share their study designs
% and results for others to reproduce, replicate, or adapt to study new
% questions in the future.

% While often effective, this approach bears two notable drawbacks.
% First, it can add substantially to the technical knowledge, computing
% resources, and initial setup needed to run or share the actual code of
% interest.  For example, sharing code for an analysis or tutorial that
% relies on a particular Docker image to run properly would of course
% necessitate writing and distributing extra configuration files and
% setup instructions.  But far more burdensomely, it also requires that
% anyone who may want to run the code (in addition to the author seeking
% to share it!) first be able to install and navigate additional
% software that is likely far more complex and resource-intensive than
% the actual analysis or tutorial code it facilitates running.  This can
% introduce a need for both a degree of computer literacy and
% computational resources that may not be universally accessible,
% particularly to students or other early-career scientists hoping to
% learn from publicly available tutorials.  These added prerequisites
% clash with the simplicity and accessibility that have contributed to
% Python's popularity, and can create significant barriers to both
% contributing to and taking advantage of open science resources.

% Second, while many existing tools allow users to initially populate a
% Python environment with a fixed set of packages and package versions
% (e.g., from a \texttt{requirements.txt}, \texttt{pyproject.toml},
% \texttt{environment.yml}, \texttt{Pipfile}, \texttt{Dockerfile RUN}
% instruction, etc.), few, if any, ensure that these specified
% requirements \textit{remain} satisfied after they are first installed.
% The ability to modify an environment after its creation is useful in
% many cases (e.g., to install additional software, when
% needed). However, this also makes it easy to inadvertently alter
% existing packages, potentially leading to subtle issues with code that
% relies on them.  For instance, suppose a researcher has implemented a
% series of analyses using version 1.0 of ``Package \textit{X}," and
% decides to perform an additional analysis that requires installing
% ``Package \textit{Y}."  If Package \textit{Y} depends on version 0.9
% of Package \textit{X}, then Package \textit{X} will be downgraded to
% accommodate this new requirement, potentially altering or breaking
% previous analyses when they are rerun later, either by the researcher
% or someone with whom they've shared their code.  Further, if some
% analyses require Package \textit{Y} while others rely on features of
% Package \textit{X} not implemented until version 1.0, it's unclear
% which version the researcher should install in their environment.

% %The \texttt{davos} package provides a novel, Python-native framework for creating, sharing, and running reproducible workflows that was designed to address each of these issues.
% The \texttt{davos} package provides a novel, Python-native framework
% for creating reproducible workflows that was designed to address each
% of these issues. \texttt{davos} allows users to specify dependencies
% directly within the code that...

% \comment {
% POINTS TO MAKE/THINGS TO SAY HERE TO WRAP UP (or maybe save some for impact section?):
% - davos simplifies the process of writing, sharing, and running reproducible code (eliminates need for additional complex env management tools, syntax is simple, packages are just installed at runtime without additional setup beforehand, etc.)...
% - ...it also provides stability/additional protections against accidental changes in ways other tools don't (i.e., addresses issue 2 above) by verifying (and installing if necessary) requirements each time code is run
% - ... while simultaneously enabling new package management patterns (using multiple versions versions of package in same notebook, etc.)
% - it's lightweight and intuitive (unlike big/complex env management tools like docker, etc.)
% - "davos was specifically designed for use with Jupyter notebook environments, and can be used either on its own or in tandem with another environment management tool"
% - "\texttt{davos} was designed specifically for use in Jupyter notebook (formerly called IPython notebook \cite{PereGran07}) environments, and supports Jupyter notebooks \cite{KluyEtal16}, JupyterLab \cite{GranGrou16}, Google Colaboratory, Binder \cite{RagaWill18}, and IDE-based notebook editors."

% - maybe don't mention smuggle statement \& onion comment explicitly in this paragraph since they haven't been defined/explained yet, or if we do, then add some sort of "see \_\_\_ below" with \\ref to relevant subsection
% }

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{snippets/snippet1.pdf}
\caption{\small \textbf{FILL THIS IN...}}
\label{fig:snippet1}
\end{figure}
